Constructing a universe for the setoid model

### 2020-07-06 paper.tex
apply lipics style

\documentclass{easychair}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{tikz}
\usepackage{bbm}
\usepackage[bw]{agda}
\usetikzlibrary{cd}

\newcommand{\mlttp}{\textsf{MLTT}^{\mProp}}
\newcommand{\refl}{\textsf{refl}}
\newcommand{\GG}{\Gamma}
\newcommand{\DD}{\Delta}
\newcommand{\GD}{\Delta}
\newcommand{\R}{\textsf{R}}
\newcommand{\Id}{\textsf{Id}}
\newcommand{\coe}{\textsf{coe}}
\newcommand{\coh}{\textsf{coh}}
\newcommand{\J}{\textsf{J}}
\newcommand{\tyj}[2]{#2 : \Ty\ #1}
\newcommand{\tmj}[3]{#2 : \Tm\ #1\ #3}
\newcommand{\subj}[3]{#1 : \mathrm{Sub}\ #2\ #3}
\newcommand{\setoidU}{\mathcal{U}}
\newcommand{\ad}[1]{\AgdaFunction{#1}}
\newcommand{\Set}{\textsf{Set}}
\newcommand{\mType}{\mathbf{Type}}
\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\mProp}{\mathbf{Prop}}
\newcommand{\Setoid}{\textsf{Setoid}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Sub}{\mathrm{Sub}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\textsf{U}}
\newcommand{\El}{\textsf{El}}
\newcommand{\Uty}{\mathds{U}}
\newcommand{\EL}{\mathds{E}l}

@@

% \usepackage{amsmath}
% \usepackage{amssymb}
% %\usepackage{mathpartir}

%\usepackage{natbib}
%\setcitestyle{round}
% Note that the \doi command from the doi package doesn't enable the
% same kinds of line breaks as the command below.
\newcommand{\doi}[1]{doi:\href{http://doi.org/#1}{%
    \urlstyle{same}\nolinkurl{#1}}}

% TODO notes.
\usepackage[textsize=small]{todonotes}
\setlength{\marginparwidth}{2cm}
\newcommand{\txatodo}[1]{\todo[fancyline,color=green!20]{#1}{}}

\title{Constructing a universe for the setoid model}
\titlerunning{Constructing a universe for the setoid model}
\authorrunning{Altenkirch, Boulier, Kaposi, Sattler, Sestini}

\author{
  Thorsten Altenkirch \inst{1}\thanks{Supported by USAF grant FA9550-16-1-0029.}
\and
  Simon Boulier
\and
  Ambrus Kaposi \inst{2}
\and
  Christian Sattler\inst{3}\thanks{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
\and
  Filippo Sestini\inst{1}
}

\institute{
  School of Computer Science, University of Nottingham, UK\\
  \email{\{psztxa,pszcs1,psxfs5\}@nottingham.ac.uk}
  \and
  E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary\\
  \email{akaposi@inf.elte.hu}
  \and
  Chalmers University of Technology, Gothenburg, Sweden
  \email{sattler@chalmers.se}
}

@@

@@

\paragraph{Contributions}

@@

\paragraph{Structure of the paper}

@@

The setoid model can be framed categorically as a category with families (CwF,
\cite{internal-tt}) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:

\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG\ \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}

In our presentation of the setoid model, contexts are given by setoids, that is,
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.

\[
\infer{
  \begin{aligned}
    & | \GG | : \mType \\
    & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
    & \textsf{refl}\ \GG : (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
    & \textsf{sym} \ \GG : \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_0 \\
    & \textsf{trans} \ \GG : \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
    \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_2
    \to \GG^\sim\,\gamma_0\,\gamma_2
  \end{aligned}
}{
  \GG : \Con
}
\]

Types in a context $\GG$ are given by displayed setoids over $\GG$
with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$.

\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & \textsf{refl*} : \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
    & \textsf{sym*} : \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
    \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
    & \textsf{trans*} : \forall\{\gamma_0\,\gamma_1\,\gamma_2\,a_0\,a_1\,a_2\}\{p_0 : \GG^\sim\ \gamma_0\ \gamma_1\}\{p_1 : \GG^\sim\ \gamma_1\ \gamma_2\}\\
    & \ \qquad \to A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
    & \textsf{coe} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \textsf{coh} : \forall\{\gamma_0\ \gamma_1 : | \GG |\} \ (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

@@

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
~
\\
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | \sigma | : | \GG | \to | \GD | \\
    & \sigma^\sim : \forall\{\rho_0\, \rho_1\} \to \GG^\sim\, \rho_0\, \rho_1 \\
    & \quad \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : \forall\{\gamma_0\, \gamma_1\}
    (p : \GG^\sim\ \gamma_0\ \gamma_1) \\
    & \quad \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\
~
\\
We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), including function extensionality and a universe of strict
propositions satisfying propositional extensionality \cite{setoid99}. Note that
we don't need identity types or inductive types (W-types) for this.

@@

We have a universe of propositions $\Prop$ defined as follows:

\[
\infer{
  \Prop : \Ty\ \GG
}{
  \GG : \Con
}
\qquad
\infer{
  \underline{P} : \Ty\ \GG
}{
  P : \Tm\ \GG\ \Prop
}
\qquad
\infer{
  u \equiv v
}{
  u : \Tm\ \GG\ \underline{P} && v : \Tm\ \GG\ \underline{P}
}
\]

Equality type constructors for contexts and dependent types internalize the idea
that every context and type comes equipped with a setoid equivalence
relation. Note that $\mProp$ is the universe of the metatheory while $\Prop$ is
the internal one. As in the model, equality for dependent types is indexed over context
equality.

\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \mProp
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ \rho_{01}\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad \rho_{01} : \GG^{\sim}\ \rho_0\ \rho_1
  }
}
\]

We have rules witnessing that these are indeed equivalence relations. We only
recall reflexivity:

\[
\infer{
  \textsf{R}\ \rho : \GG^\sim\ \rho\ \rho
}{
  \rho : \Sub\ \GD\ \GG
}
\qquad
\infer{
  \textsf{R}\ a : \Tm\ \GG\ \underline{A^\sim\ (\textsf{R}\ \rho)\ a\ a}
}{
  A : \Ty\ \GG
  \qquad
  \rho : \Sub\ \GD\ \GG
  \qquad
  a : \Tm\ \GD\ A[\rho]
}
\]

In addition, we also have rules representing the fact that every construction in
SeTT respects setoid equality, so that we can transport along any such equality:

    \[
    \infer{
      \deduce{
        \tmj{\GD}{\coh_A\ p\ a}{\underline{A^\sim\ p\ a\ (\coe_A\ p\ a)}}
      }{
        \tmj{\GD}{\coe_A\ p\ a}{A[\rho_1]}
      }
    }{
      \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
      p : \GG^\sim\ \rho_0\ \rho_1\qquad
      \tmj{\GD}{a}{A[\rho_0]}
    }
    \]

@@

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.

\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t :\equiv
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]

@@

As an example, consider the following definition of a small universe with
booleans and $\Pi$ types.

@@

In our case, we can encode $\U$ as an inductive type $\ad{in-U}$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\ad{in-U}$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\ad{in-U}$ quite expectedly reflects the
inductive structure of codes.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \mType \to \mType_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \mType\} \{B : A \to \mType\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}

$\U$ and $\El$ can be given as follows:
\[
\U = \Sigma (A : \mType)\ (\textsf{in-U}\ A) \qquad \El = \pi_1
\]

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations:
%

\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\
~
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$:
%

\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\
~
\\

@@

This gives a complete definition of a universe of setoids, which is, however,
inductive-recursive. Moreover, the kind of recursion involved in this definition
is particularly complex, and not obviously reducible to well-understood notions
of induction-recursion like the one described in \cite{dybjer-setzer}. In any
case, we would like to avoid extending the metatheory with any forms of
induction-recursion, given our goal to keep the metatheory as small and
essential as possible.

@@

\bibliographystyle{alpha}
\bibliography{bibliography}

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2019}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathpartir}
\usepackage{dsfont}
\usepackage{proof}
\usepackage{bbm}
%\usepackage[bw]{agda}

\bibliographystyle{plainurl}% the mandatory bibstyle

\newcommand{\mlttp}{\textsf{MLTT}^{\mProp}}
\newcommand{\refl}{\textsf{refl}}
\newcommand{\GG}{\Gamma}
\newcommand{\DD}{\Delta}
\newcommand{\GD}{\Delta}
\newcommand{\R}{\textsf{R}}
\newcommand{\Id}{\textsf{Id}}
\newcommand{\coe}{\textsf{coe}}
\newcommand{\coh}{\textsf{coh}}
\newcommand{\J}{\textsf{J}}
\newcommand{\tyj}[2]{#2 : \Ty\ #1}
\newcommand{\tmj}[3]{#2 : \Tm\ #1\ #3}
\newcommand{\subj}[3]{#1 : \mathrm{Sub}\ #2\ #3}
\newcommand{\setoidU}{\mathcal{U}}
% \newcommand{\ad}[1]{\AgdaFunction{#1}}
\newcommand{\inU}{\mathsf{inU}}
\newcommand{\inUU}{\mathsf{inU{\sim}}}
\newcommand{\Set}{\textsf{Set}}
\newcommand{\mType}{\mathbf{Type}}
\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\mProp}{\mathbf{Prop}}
\newcommand{\Setoid}{\textsf{Setoid}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Sub}{\mathrm{Sub}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\textsf{U}}
\newcommand{\El}{\textsf{El}}
\newcommand{\Uty}{\mathds{U}}
\newcommand{\EL}{\mathds{E}l}

@@

% TODO notes.
%\usepackage[textsize=small]{todonotes}
%\setlength{\marginparwidth}{2cm}
%\newcommand{\txatodo}[1]{\todo[fancyline,color=green!20]{#1}{}}

\title{Constructing a universe for the setoid model}
%\titlerunning{Constructing a universe for the setoid model}

\author{Thorsten Altenkirch}{School of Computer Science, University of Nottingham, UK}{psztxa@nottingham.ac.uk}{orcid}{Supported by USAF grant FA9550-16-1-0029.}
\author{Simon Boulier}{Inria, Nantes, France}{simon.boulier@inria.fr}{}{}
\author{Ambrus Kaposi}{E{\"o}tv{\"o}s Lor{\'a}nd University, Budapest, Hungary}{akaposi@inf.elte.hu}{orcid}{Supported by the National Research, Development and Innovation Fund of Hungary, financed under the Thematic Excellence Programme funding scheme, Project no. ED\_18-1-2019-0030 and the Bolyai Fellowship of the Hungarian Academy of Sciences, Project no.\ BO/00659/19/3.}
\author{Christian Sattler}{Chalmers University of Technology, Gothenburg, Sweden}{sattler@chalmers.se}{}{Supported by USAF grant FA9550-16-1-0029 and Swedish Research Council grant 2019-03765.}
\author{Filippo Sestini}{School of Computer Science, University of Nottingham, UK}{Filippo.Sestini@nottingham.ac.uk}{}{}

\authorrunning{T. Altenkirch and S. Boulier and A. Kaposi and C. Sattler and F. Sestini}

\Copyright{Thorsten Altenkirch and Simon Boulier and Ambrus Kaposi and Christian Sattler and Filippo Sestini} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[500]{Theory of computation~Logic~Type theory} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{type theory, function extensionality, univalence, setoid model, induction-recursion, induction-induction} %TODO mandatory; please add comma-separated list of keywords

%\category{} %optional, e.g. invited paper

%\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
The setoid model of type theory can be used to show the consistency of
intensional type theory with certain extensionality principles, like function
extensionality. In addition, Altenkirch's formulation of such model justifies a
universe of strict (definitionally-proof irrelevant) propositions supporting
propositional extensionality, a limited form of univalence according to which
logically equivalent propositions are equal. The appeal of this model
construction is that it takes place in a small, intensional, type theoretic
metatheory, therefore giving a method to boostrap extensionality.

Recently, Altenkirch's setoid model has been turned into a formal system, under
the name of Setoid Type Theory (SeTT). SeTT can be presented as an extension of
intensional Martin-Loef type theory with syntactic constructs that internalize
setoid reasoning. These give full access to the extensionality principles that
hold in the setoid model, while retaining the convenience of an intensional
theory.

As currently defined, SeTT is already a rich theory. However, its introspection
capabilities are currently lacking, since it only includes (univalent) universes
for propositions. We would like to extend SeTT with a universe of setoids
internalizing the notion of type beyond propositions. This amounts to extending
the setoid model with a semantics for such a universe.

In this paper, we present the construction of a (non-univalent) universe of
setoids within the setoid model. The universe is given in steps, first as an
inductive-recursive definition, which is then translated to an
inductive-inductive definition, and finally to an inductive family. Every step
is obtained from the previous by applying known reduction methods, although in a
novel way. Additionally, our construction shows the setoid model with the
additional encoding of the universe can still be defined in a relatively small
metatheory, that however must include a universe of strict propositions, indexed
W types, and a proof-irrelevant identity type with a strong transport rule. All
the mathematical developments described in the paper have been formalized in the
Agda proof-assistant.
\end{abstract}

@@

\paragraph*{Contributions}

@@

\paragraph*{Structure of the paper}

@@

The setoid model can be framed categorically as a category with families (CwF,
\cite{internal-tt}) with extra structure for the various type and term
formers. The core structure of a CwF can be given as the following signature:
\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG\ \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}

In our presentation of the setoid model, contexts are given by setoids, that is,
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.
\[
\infer{
  \begin{aligned}
    & | \GG | : \mType \\
    & \GG^\sim : | \GG | \to | \GG | \to \mProp \\
    & \textsf{refl}\ \GG : (\gamma : | \GG |) \to \GG^\sim\,\gamma\,\gamma \\
    & \textsf{sym} \ \GG : \forall\{\gamma_0\ \gamma_1\} \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_0 \\
    & \textsf{trans} \ \GG : \forall\{\gamma_0\ \gamma_1\ \gamma_2\}
    \to \GG^\sim\,\gamma_0\,\gamma_1
    \to \GG^\sim\,\gamma_1\,\gamma_2
    \to \GG^\sim\,\gamma_0\,\gamma_2
  \end{aligned}
}{
  \GG : \Con
}
\]

Types in a context $\GG$ are given by displayed setoids over $\GG$
with a fibration condition given by $\textsf{coe}$, $\textsf{coh}$. In
the following, we somtimes omit implicit quantifications such as the $\forall\{\gamma_0\ \gamma_1\}$ in the type of $\textsf{sym}\,\GG$.
\[
\infer{
  \begin{aligned}
    & | A | : | \GG | \to \mType \\
    & A^\sim : \{\gamma_0\ \gamma_1 : | \GG |\}
    \to \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \to \mProp \\
    & \textsf{refl*} : \{\gamma : | \GG | \} (a : | A | \gamma) \to A^\sim\ (\textsf{refl}\ \GG\ \gamma)\ a\ a \\
    & \textsf{sym*} : \forall\{\gamma_0\,\gamma_1\,a_0\,a_1\}\{p : \GG^\sim\ \gamma_0\ \gamma_1\}
    \to A^\sim\ p\ a_0\ a_1 \to A^\sim\ (\textsf{sym}\ \GG\ p)\ a_1\ a_0 \\
    & \textsf{trans*} : A^\sim\ p_0\ a_0\ a_1\to A^\sim\ p_1\ a_1\ a_2\to A^\sim\ (\textsf{trans}\,\GG\,p_0\,p_1)\ a_0\ a_2\\
    & \textsf{coe} : \GG^\sim\ \gamma_0\ \gamma_1 \to | A | \gamma_0 \to | A | \gamma_1 \\
    & \textsf{coh} : (p : \GG^\sim\ \gamma_0\ \gamma_1) (a : | A | \gamma_0)
    \to A^\sim\,p\,a\,(\textsf{coe}\,A\,p\,a) \\
  \end{aligned}
}{
  A : \Ty\ \GG
}
\]

@@

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | \sigma | : | \GG | \to | \GD | \\
    & \sigma^\sim : \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    & | t | : (\gamma : | \GG |) \to | A | \, \gamma \\
    & t^\sim : (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\

We can show that the setoid model validates the usual basic type formers ($\Pi,
\Sigma$, etc.), function extensionality and a universe of strict
propositions ith propositional extensionality \cite{setoid99}. Note that
we don't need identity types or inductive types (W-types) for this.

@@

We have a universe of propositions $\Prop$ defined as follows:
\[
\infer{
  \Prop : \Ty\ \GG
}{
  \GG : \Con
}
\qquad
\infer{
  \underline{P} : \Ty\ \GG
}{
  P : \Tm\ \GG\ \Prop
}
\qquad
\infer{
  u \equiv v\vphantom{\Prop}
}{
  u : \Tm\ \GG\ \underline{P} && v : \Tm\ \GG\ \underline{P}
}
\]

Equality type constructors for contexts and dependent types internalize the idea
that every context and type comes equipped with a setoid equivalence
relation. Note that $\mProp$ is the universe of the metatheory while $\Prop$ is
the internal one. As in the model, equality for dependent types is indexed over context
equality.
\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \mProp
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ \rho_{01}\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad \rho_{01} : \GG^{\sim}\ \rho_0\ \rho_1
  }
}
\]

We have rules witnessing that these are indeed equivalence relations. We only
recall reflexivity:
\[
\infer{
  \textsf{R}\ \rho : \GG^\sim\ \rho\ \rho
}{
  \rho : \Sub\ \GD\ \GG
}
\qquad
\infer{
  \textsf{R}\ a : \Tm\ \GG\ \underline{A^\sim\ (\textsf{R}\ \rho)\ a\ a}
}{
  A : \Ty\ \GG
  \qquad
  \rho : \Sub\ \GD\ \GG
  \qquad
  a : \Tm\ \GD\ A[\rho]
}
\]

In addition, we also have rules representing the fact that every construction in
SeTT respects setoid equality, so that we can transport along any such equality:
    \[
    \infer{
      \deduce{
        \tmj{\GD}{\coh_A\ p\ a}{\underline{A^\sim\ p\ a\ (\coe_A\ p\ a)}}
      }{
        \tmj{\GD}{\coe_A\ p\ a}{A[\rho_1]}
      }
    }{
      \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
      p : \GG^\sim\ \rho_0\ \rho_1\qquad
      \tmj{\GD}{a}{A[\rho_0]}
    }
    \]

@@

We can easily recover the usual Martin-L\"of identity type from setoid equality,
with transport implemented via coercion.
\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]
\[
\infer{
  \textsf{transp}\ P\ p\ t :\equiv
    \coe\ P\ (\textsf{R}\ \textsf{id} , p)\ t : \Tm\ \GG\ P[a_1]
}{
  P : \Ty\ (\GG . A)
  \qquad
  p : \Tm\ \GG\ (\Id\ A\ a_0\ a_1)
  \qquad
  t : \Tm\ \GG\ P[a_0]
}
\]

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$.

@@

In our case, we can encode $\U$ as an inductive type $\inU$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\inU$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\inU$ quite expectedly reflects the
inductive structure of codes.
%
\begin{alignat*}{10}
  & \textsf{data}\ \inU && : \mType \to \mType_1 \\
  & \quad \textsf{inBool} && : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
\end{alignat*}

$\U$ and $\El$ can be given by $\U :\equiv \Sigma (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\
%
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data} \ \setoidU : \mType_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{refl}_\setoidU : (A : \setoidU) \to \equ{A}{A} \\
  & \textsf{sym}_\setoidU : \equ{A}{B} \to \equ{B}{A} \\
  & \textsf{trans}_\setoidU : \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}
\\

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
%
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \Uty | :\equiv \lambda\, \gamma . \, \setoidU \\
  & \Uty^\sim :\equiv \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  & \textsf{refl}\ \Uty :\equiv \reflu \\
  & ... \\
  & \textsf{coe}\ \Uty :\equiv \lambda\,p\,a . \, a \\
  & \textsf{coh}\ \Uty :\equiv \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & | \EL\ A | :\equiv \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  & (\EL\ A)^\sim :\equiv \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  & \textsf{refl}\ (\EL\ A) :\equiv \reflel \\
  & ... \\
  & \textsf{coe}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  & \textsf{coh}\ (\EL\ A) :\equiv \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}
\\

@@

This gives a complete definition of a universe of setoids, which is, however,
inductive-recursive. Moreover, the kind of recursion involved in this definition
is particularly complex, and not obviously reducible to well-understood notions
of induction-recursion like the one described in \cite{dybjer-setzer}. In any
case, we would like to avoid extending the metatheory with any form of
induction-recursion in order to keep the metatheory as small and
essential as possible.

@@

\bibliography{bibliography}
